The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

### 2024-03-20 article/paper_drought_Chile2023.qmd
update article

@@

For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017]. Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020]. One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022]. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation. SPEI allows for analysis of the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021]. @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought from short- to long-term scales and consider variables from the earth's water balance on their own (e.g., precipitation, AED, soil moisture). Using this information, we can advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought. [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. There is limited understanding of the correlation between drought and the environment in Chile. Therefore, it is crucial to have a deeper understanding of how meteorological and soil moisture droughts impact the dynamics of the environment in order to make informed decisions on how to adapt.

@@

## Trend of short- to long-term drought

@@

We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses. To achieve this, we used the persistent mask of land cover macroclasses. This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year.

We examine the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity. The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity. We want to address three main questions: Which of the drought variables—supply, demand, or soil moisture—most helps to explain the changes in plant productivity? Which of the short-term or long-term time scales of the drought variable has a bigger effect on vegetation productivity in Chile? And finally, how strong is the relationship between the variables and the drought index? Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five land cover types: forest, cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months with zcNDVI. We used a method similar to that used by @Meroni2017 which compared the SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation). A pixel-to-pixel linear correlation analysis was performed for each index within the persistent mask of land cover macroclasses. To begin, the Pearson coefficient of correlation is computed for each of the six time scales. A time scale is identified as the one that attains the highest correlation (p < 0.05). Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted. As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales and drought index that reached the maximum correlation, and 2) having the magnitude of the correlation obtained by the drought index at the time scales.

@@

For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023]. For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019]. For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023]. For managing vectorial data, we used `{sf}` [@Pebesma2018]. For the calculation of AED, we used `{SPEI}` [@Bergueria2023]. For mapping, we use {tmap} [@Tennekes2018]. For data analysis, the suite {tidyverse} [@Wickham2019] was used. For the random forest modeling, we used the {tidymodels}[@Kuhn2020] and {ranger}[@Wright2017] packages.

@@

In @fig-trendDIMacro, the averaged aggregation per macrozone, drought index, and time scale are shown. The macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales. Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile. At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI. For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI. "Norte Grande" has the highest trend at 36 months for EDDI (0.042 per decade), and "Centro" has the lowest for SPI and SPEI. In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones. Contrary to the other macrozones, “Austral” showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.

### Interaction of land cover and drought

## Land cover change

@@

From the trend analysis in Table \ref{tab-landcoverTrend}, we can indicate that the "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$ and a reduction in the class savanna of 70 $km^2 yr^{-1}$. In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).

@@

In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a). In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.027 for grassland and 0.032 for shrubland.  In the "Norte Chico," savanna has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and grassland -0.037. In "Centro," shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade. This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].

The temporal variation within the macrozones is showed in [@fig-zcNDVI_var]b). There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.  Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro". The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity across Chile.

@@

@fig-corTimeScale is a map that shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and plant productivity over time. The spatial variation of time scales reached per index is mostly for time scales above 12 months. In the case of SSI, the predominant scales are 6 and 12 months. For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral" increases. In @fig-corPerson, the map of Pearson correlation values is shown. The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6. Following a similar spatial pattern as EDDI. The SSI showed to be the index that has a major spatial extension with a higher correlation. It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."

@@

## Drought trend and attribution to land cover

@Vicente-Serrano2022, in a study at the global scale of drought trends, indicate that there have not been significant trends in meteorological drought since 1950. Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction. @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity."_ In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover, to avoid misleading conclusions that could be related to the increase in water demand due to land cover change. Our results show a contrasting perspective. There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the mega drought [@Garreaud2017]. However, a rise in irrigated land doesn't seem to have an impact on this hydrological drought. Despite using the persistance mask for vegetation's trend analysis, cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types. These changes are associated with a decrease in water demand from vegetation. Nonetheless, we used the persistant land cover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption and consequently increase water demand. But this effect should be minor compared to the results from land cover change at this scale of analysis.

On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales, which is evidence of the hydrological drought. We say that what happened in central Chile goes against what has been found on a global scale [@Vicente-Serrano2022;@Kogan2020]. This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand by vegetation like irrigated crops. Finally, north-central Chile has experienced a decline in vegetation productivity across all macroclasses, which is primarily attributable to variations in water supply and soil moisture. An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.

## Land cover types and their impact by drought

We discovered that croplands, savannas, and shrubland are the most susceptible to climatic changes and are most affected by the 12-month soil moisture deficit. In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI). They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6). In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010]. Short-term precipitation deficits impact soil water, and thus less water is available for plant growth. However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12. Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater. In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared. What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types. Also, drought episodes have been less frequent and intense. The drought episodes have had a lower impact on water availability for vegetation.

According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality. However, we found that forest is the type of land cover macroclass less affected by variation in drought indices, being the most resistant land cover class to drought. Supporting this is @Fathi-Taperasht2022], who assert that Indian forests are the most drought-resistant and recover rapidly. Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance. Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 test the resistance of ecosystems and find that ecosystems with more forests are better able to handle severe droughts than croplands. They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].

In contrast to what we obtained, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down. This could mean that the natural dynamics of their forests have changed. They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought). Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity compared to the other land cover classes. Nonetheless, if we want to go deep in our analysis, we should use earth observation data that is able to capture a higher level of detail. For example, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer. Then, to use remote sensing to look at how a certain type of forest (like sclerophyllous forest) changes in response to drought on a local level, we should use operational products with higher spatial resolutions, like those from Landsat or Sentinel. This will let us do a more thorough analysis.

## Soil moisture, vegetation productivity, and agricultural drought.

>>>>

  - vegetation productivity

@@

For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017]. Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020]. One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022]. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation. SPEI allows for analysis of the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021]. @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from the earth's water balance (e.g., precipitation, AED, soil moisture).  The variation in climate variables impacts vegetation development, and unfavorable conditions such as low precipitation and high temperatures usually generate a decrease in vegetation productivity. To monitor the response of vegetation, the common practice is to use satellite data. The Normalized Difference Vegetation Index (NDVI) has been widely used as a proxy for biomass production [@Camps-Valls2021;@Paruelo2016;@Helman2014]. For Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in response to drought.  Using this information, we can advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought. [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. The association between drought and the environment in Chile is not well comprehended. Hence, it is imperative to acquire a more profound comprehension of the manner in which climatic and soil moisture droughts influence environmental dynamics, in order to make well-informed decisions on adaptation strategies.

@@

## Short- to long-term drought

@@

We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses. To achieve this, we used the persistent mask of land cover macroclasses. This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year. We used the zcNDVI as a proxy of vegetation productivity. In Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in relation to climate.

We examine the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity. The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity. We want to address three main questions: Which of the drought variables—supply, demand, or soil moisture—most helps to explain the changes in plant productivity? How do the short- to long-term time scales of the drought variable affect vegetation productivity in Chile? And finally, how strong is the relationship between the variables and the drought index? Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five land cover types: forest, cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months with zcNDVI. We used a method similar to that used by @Meroni2017 which compared the SPI time-scales with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation). We performed a pixel-to-pixel linear correlation analysis for each index within the persistent mask of land cover macroclasses. We first compute the Pearson coefficient of correlation for each of the six time scales. A time scale is identified as the one that attains the highest correlation (p < 0.05). We then extracted the Pearson correlation coefficient corresponding to the time scales where the value peaked. As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales and drought index that reached the maximum correlation, and 2) having the magnitude of the correlation obtained by the drought index at the time scales.

@@

For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023]. For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019]. For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023]. For managing vectorial data, we used `{sf}` [@Pebesma2018]. For the calculation of AED, we used `{SPEI}` [@Bergueria2023]. For mapping, we use {tmap} [@Tennekes2018]. For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used. For the random forest modeling, we used the {tidymodels}[@Kuhn2020] and {ranger}[@Wright2017] packages.

@@

The @fig-trendDIMacro displays the averaged aggregation per macrozone, the drought index, and the timescale. The macrozones that reached the lowest trend for SPI, SPEI, and SSI are "Norte Chico" and "Centro," where the indices also decrease at longer time scales. Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile. At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI. For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI. "Norte Grande" has the highest trend at 36 months for EDDI (0.042 per decade), and "Centro" has the lowest for SPI and SPEI. In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones. Contrary to the other macrozones, “Austral” showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.

@@

From the trend analysis in Table \ref{tab-landcoverTrend}, we can indicate that the "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$ and a reduction in the class savanna of 70 $km^2 yr^{-1}$. In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," seemingly replacing the savanna lost (Table \ref{tab-landcoverTrend}).

@@

The temporal variation within the macrozones is shown in [@fig-zcNDVI_var]b). There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.  Vegetation reached its lowest values for 2019-2022, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro". The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively, per decade (@fig-zcNDVI_var).

In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a). In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.027 for grassland and 0.032 for shrubland.  In the "Norte Chico," savanna has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and grassland -0.037. In "Centro," shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade. This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning.

<!-- Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity across Chile. -->

@@

@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and zcNDVI over time scales (1, 3, 6, 12, 24, and 36 months). The spatial variation of time scales reached per index is mostly for time scales above 12 months. In the case of SSI, the predominant scales are 6 and 12 months. For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral," where they increase. In @fig-corPerson, the map of Pearson correlation values (r) is shown. The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6. Following a similar spatial pattern as EDDI but with an opposite sign. The SSI showed to be the index that has a major spatial extension with a higher correlation. It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."

@@

## Drought trend and attribution to land cover change

@Vicente-Serrano2022, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950. Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction. @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that "since 1980, the Earth warming has not changed the drought area or intensity." In our study, we took into account the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change. Our results show a contrasting perspective. The SPI, SPEI, and SSI (water supply) showed a decrease in trends, except for the southern part, and an increase in EDDI (water demand). The trend, positive or negative, was stronger as the time scales increased. Trends in the long term (e.g., 36 months) are evidence of how human-induced climate change is affecting Chile, which seems to be due to an intense hydrological drought resulting from the persistence of the precipitation deficit. We found that there has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for the north-central part of the country, which has been extreme between 2020 and 2022 and has impacted natural and cultivated land. Additionally, we demonstrated that the drought, primarily due to an increase in AED, accounts for about 30% of the changes in land cover types (excluding croplands). These changes are associated with a decrease in water demand from vegetation. Moreover, the most water-demanding type, cropland, showed a decrease in the north-central region, while barren land showed an increase. The north-central part of the country primarily experienced these changes due to a higher increase in AED. Thus, we have evidence of a significant decline in water supply and an increase in AED for the north-central part of Chile, which show to be the most relevant variables for drought conditions. Some questions arise regarding what is occurring with the cultivated land. We used the unchanged land cover to ensure that an increase in surface area is not considered in the trend analysis. For croplands, it could happen that some areas have changed the types of crops for others with higher water demand, which consequently increases water demand. However, this effect should be minor compared to the decrease in water supply and increase in water demand at this scale of analysis.

This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops). North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture. An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.

## Impact of short- to long-term droughts on land cover types

We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile. In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI). They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6). In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010]. Short-term precipitation deficits impact soil water, and thus less water is available for plant growth. However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12. Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater. In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared. The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense. The drought episodes have had a lower impact on water availability for vegetation.

According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality. However, we discovered that forests, as the most resilient land cover class to drought, experience less variation in drought indices. Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly. Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance. Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands. They attribute the difference to a deeper rooting depth for trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023]. In contrast, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down. This could mean that the natural dynamics of their forests have changed. They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).

Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity compared to the other land cover classes. Nonetheless, if we want to go deep in our analysis, we should use earth observation data that is able to capture a higher level of detail. For example, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer. Then, to use remote sensing to look at how a certain type of forest (like sclerophyllous forest) changes in response to drought on a local level, we should use operational products with higher spatial resolutions, like those from Landsat or Sentinel. This will let us do a more thorough analysis.

## Vegetation productivity and drought.
